---
title: "An Introduction to zitools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Installation

You can install the development version of zitools from GitHub with:

```{r, eval = FALSE}
devtools::install_github("kreutz-lab/zitools")
```

# Introduction 

This vignette provides an introductory example on how to work with the 'zitools' package, which implements a weighting strategy based on a fitted zero inflated mixture model.

Step one: load required packages

```{r}
library(zitools)
```

## ziMain function 

zitools is a ... tool to analyse zero inflated count data, especially microbiome data. As per default the counts are estimated given the predictor variables sample (columns) and feature(rows) when fitting a zero inflated poisson or zero inflated negative binomial model.

```{r}
data(mtx)
Zi <- ziMain(mtx)


```

### Plots 

```{r}
boxplot(Zi)
boxplot(Zi@countmatrix)

heatmap(Zi, Rowv=NULL)
```
